---

title:  "Project 2 — Data Tidying"
author: "Shawn Ganz"
date:   today
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  echo:    true
  warning: false
---

## Approach

I'll use the following data:


- https://brightspace.cuny.edu/d2l/le/1154241/discussions/threads/4168991/View
  - df_1
- https://brightspace.cuny.edu/d2l/le/1154241/discussions/threads/4206056/View
  - df_2
- https://brightspace.cuny.edu/d2l/le/1154241/discussions/threads/4224007/View
  - df_3
  - Wasn't wide, so I have to make it wide

```{r}
df <- data.frame(
  Quarter = c("Q1","Q2","Q3","Q4"),
  Lakers = c(28,32,25,30),
  Celtics = c(31,29,33,28),
  Warriors = c(25,27,35,30)
)

write.csv(df, "data_03.csv", row.names = FALSE)
```

put csv's onto github
read csv in project

- `url -> "github_raw_link"`
- `library(tidyverse)`
- `df <- read_csv("url")`

**Analysis for df_1**

Fix the brackets, there are too many. 

- Summary table
  - religion
  - income bracket 
  - n values
  
- ggplot
  - x = income brackets
  - y = n 
  - fill = religion

- Description
  - Talk about which ones has the most per bracket
  - Maybe even 
  
**Analysis for df_2**

The only real wide portion of the dataset is the months. So, make that into months using `pivot_longer()` and try and fix the craziness of screeningscores and `pivot_wider()` PHQ and GAD. Limit selection to a couple of rows and then see if there's anything to analyze. 

Could show off age, month, and values.

- Summary table
  - group age, month
  - summarize
    - values
  
- ggplot
  - x = age
  - y = values 
  - cols
  
- Description
  - findings

**Analysis for df_3**

Simple one, will just `pivot_longer()` team names into new column, values to score. 

- Summary table
  - entire dataset
  
- ggplot
  - x = quarter
  - y = scores
  - fill = team
  - cols
  
- Description
  - review the chart

## Codebase

```{r}
library(tidyverse)
url_1 <- "https://raw.githubusercontent.com/Siganz/CUNY_Assignments/refs/heads/main/607/project_02/data_01.csv"
url_2 <- "https://raw.githubusercontent.com/Siganz/CUNY_Assignments/refs/heads/main/607/project_02/data_02.csv"
url_3 <- "https://raw.githubusercontent.com/Siganz/CUNY_Assignments/refs/heads/main/607/project_02/data_03.csv"

df_1 <- read_csv(url_1)
df_2 <- read_csv(url_2)
df_3 <- read_csv(url_3)

#df_all <- c(df_1, df_2, df_3)
#glimpse(df_all <- c(df_1, df_2, df_3))

glimpse(df_1)
glimpse(df_2)
glimpse(df_3)
```

```{r}
df_1 |>
  mutate(
  low = rowSums(across(c(`<$10k`, `$10-20k`, `$20-30k`, `$30-40k`, `$40-50k`))),
  mid = rowSums(across(c(, `$50-75k`, `$75-100k`))),
  high = rowSums(across(c(`$100-150k`, `>$150k`)))) |>
  select(Religion, low, mid, high) |>
  pivot_longer(cols = -Religion, names_to = "income_range", values_to = "n")

df_2 |>
  select(RecordID, AgeGroup, ScreeningScores, TherapyType, Jan, Feb, Mar, Apr, May) |>
  rename(id = RecordID, scores = ScreeningScores, age = AgeGroup) |>
  pivot_longer(cols = c(Jan, Feb, Mar, Apr, May), names_to = "month", values_to = "values") |>
  mutate(scores = str_replace(scores, ",\\s*", ";"),
         scores = str_replace_all(scores, "[:\\-=]", ""),
         scores = str_replace(scores, "GAD7", ""),
         scores = str_replace(scores, "PHQ9", ""))|> 
  separate(col = scores, into = c("PHQ", "GAD7"), sep = ";") |>
  mutate(values = replace(values, values %in% c("-", "missing", "", "n/a"), NA))

df_3 |> rename(quarter = Quarter) |>
  pivot_longer(cols = -quarter, names_to = "teams", values_to = "scores") |>
  arrange(teams)
```