---
title:  "Project 2 — Data Tidying"
author: "Shawn Ganz"
date:   today
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  echo:    true
  warning: false
---

## Approach

I'll use the following data:


- [Dataset_1](https://brightspace.cuny.edu/d2l/le/1154241/discussions/threads/4168991/View)
  - df_1
- [Dataset_2](https://brightspace.cuny.edu/d2l/le/1154241/discussions/threads/4206056/View)
  - df_2
- [Dataset_3](https://brightspace.cuny.edu/d2l/le/1154241/discussions/threads/4224007/View)
  - df_3
  - Wasn't wide, so I have to make it wide

```{r}
df <- data.frame(
  Quarter = c("Q1","Q2","Q3","Q4"),
  Lakers = c(28,32,25,30),
  Celtics = c(31,29,33,28),
  Warriors = c(25,27,35,30)
)

write.csv(df, "data_03.csv", row.names = FALSE)
```

put csv's onto github
read csv in project

- `url -> "github_raw_link"`
- `library(tidyverse)`
- `df <- read_csv("url")`

**Analysis for df_1**

Fix the brackets, there are too many. 

- Summary table
  - religion
  - income bracket 
  - n values
  
- ggplot
  - x = income brackets
  - y = n 
  - fill = religion

- Description
  - Talk about which ones has the most per bracket
  - Maybe even 
  
**Analysis for df_2**

The only real wide portion of the dataset is the months. So, make that into months using `pivot_longer()` and try and fix the craziness of screeningscores and `pivot_wider()` PHQ and GAD. Limit selection to a couple of rows and then see if there's anything to analyze. 

Could show off age, month, and values.

- Summary table
  - group age, month
  - summarize
    - values
  
- ggplot
  - x = age
  - y = values 
  - cols
  
- Description
  - findings

**Analysis for df_3**

Simple one, will just `pivot_longer()` team names into new column, values to score. 

- Summary table
  - entire dataset
  
- ggplot
  - x = quarter
  - y = scores
  - fill = team
  - cols
  
- Description
  - review the chart

## Codebase

**Reading**
```{r}
library(tidyverse)
url_1 <- "https://raw.githubusercontent.com/Siganz/CUNY_Assignments/refs/heads/main/607/project_02/data_01.csv"
url_2 <- "https://raw.githubusercontent.com/Siganz/CUNY_Assignments/refs/heads/main/607/project_02/data_02.csv"
url_3 <- "https://raw.githubusercontent.com/Siganz/CUNY_Assignments/refs/heads/main/607/project_02/data_03.csv"

df_1 <- read_csv(url_1)
df_2 <- read_csv(url_2)
df_3 <- read_csv(url_3)

#df_all <- c(df_1, df_2, df_3)
#glimpse(df_all <- c(df_1, df_2, df_3))

glimpse(df_1)
glimpse(df_2)
glimpse(df_3)
```

**Cleaning**
```{r}
religion_df <- df_1 |>
  rename(religion = Religion) |>
  mutate(
  low = rowSums(across(c(`<$10k`, `$10-20k`, `$20-30k`, `$30-40k`, `$40-50k`))),
  mid = rowSums(across(c(, `$50-75k`, `$75-100k`))),
  high = rowSums(across(c(`$100-150k`, `>$150k`)))) |>
  select(religion, low, mid, high) |>
  pivot_longer(cols = -religion, names_to = "income_range", values_to = "n")
glimpse(religion_df)

medical_df <- df_2 |>
  select(RecordID, AgeGroup, ScreeningScores, TherapyType, Jan, Feb, Mar, Apr, May) |>
  rename(id = RecordID, scores = ScreeningScores, age = AgeGroup) |>
  pivot_longer(cols = c(Jan, Feb, Mar, Apr, May), names_to = "month", values_to = "values") |>
  mutate(scores = str_replace(scores, ",\\s*", ";"),
         scores = str_replace_all(scores, "[:\\-=]", ""),
         scores = str_replace(scores, "GAD7", ""),
         scores = str_replace(scores, "PHQ9", ""))|> 
  separate(col = scores, into = c("PHQ", "GAD7"), sep = ";") |>
  mutate(values = replace(values, values %in% c("-", "missing", "", "n/a"), NA)) |>
  mutate(value = as.integer(values), PHQ = as.integer(PHQ), GAD7 = as.integer(GAD7)) |>
  select(-values) |>
  mutate(TherapyType = ifelse(TherapyType == "n/a", "None", TherapyType),
         month = factor(month, levels = c("Jan", "Feb","Mar", "Apr", "May")))
glimpse(medical_df)

basketball_df <- df_3 |> rename(quarter = Quarter) |>
  pivot_longer(cols = -quarter, names_to = "teams", values_to = "scores") |>
  mutate(quarter = factor(quarter, levels = c("Q1","Q2","Q3","Q4"))) |>
  arrange(teams)
glimpse(basketball_df)
```

**Religion Summarize + ggplot**
```{r}
library(ggthemes)

sum_religion <- religion_df |> 
  mutate(income_range = factor(income_range, levels = c("low", "mid", "high"))) |>
  group_by(religion, income_range) |>
  summarize(n = sum(n))

sum_religion |>
  ggplot(aes(x = income_range, y = n, fill = religion)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Religion counts by income range",
    x = "income range",
    y = "count")

religion_df |> group_by(religion) |>
  summarize(n = sum(n)) |>
  arrange(desc(n)) |>
  slice_head(n = 3)

religion_df |> group_by(religion) |>
  summarize(n = sum(n)) |>
  arrange(n) |>
  slice_head(n = 3)

```
According to the religion data, there are more patrons that are of the low income range ($0 - $50k), than the other income ranges. One notable pattern is that Orthodox has more patrons that are high income than mid income that can be visually determined.

The religions with the highest count of patrons:
- Evangelical (7943)
- Catholic  (6724)
- Mainline Protestant (6300)

The religions with the lowest count of patrons:
- Jehovah's Witness (178)
- Hindu  (232)
- Muslim (234)

**Medical Summarize + ggplot**

```{r}
sum_medical <- medical_df |> 
  rename(therapy = TherapyType) |>
  filter(!is.na(value)) |>
  group_by(therapy, month) |>
  summarize(n = sum(value),
            mean_phq = mean(PHQ, na.rm = TRUE),
            mean_gad = mean(GAD7, na.rm = TRUE),
            .groups = "drop")

glimpse(sum_medical)

sum_medical |>
  group_by(month, therapy) |>
  ggplot(aes(month, mean_phq, color = therapy, group = therapy)) +
  geom_line() +
  geom_point() + 
  labs(title = "Mean PHQ over time",
       x = "Month",
       y = "Mean PHQ Score")

sum_medical |>
  group_by(month, therapy) |>
  ggplot(aes(month, mean_gad, color = therapy, group = therapy)) +
  geom_line() +
  geom_point() +
  labs(title = "Mean GAD over time",
       x = "Month",
       y = "Mean GAD Score")
```
We can see the highest Mean PHQ score was in April with the Group therapy, which also had the largest drop in the following month. We can see the same in the GAD ggplot, where the highest GAD score was in January with the ACT therapy, followed by the largest drop. 

**Basketball Summarize + ggplot**

```{r}

basketball_df |>
  ggplot(aes(quarter, scores, fill = teams)) +
  geom_col(position = "dodge") +
  labs(title = "Scores Per Team Per Quarter",
       x = "Quarters",
       y = "Score")

basketball_df |>
  group_by(teams) |>
  slice_max(scores, n = 1, with_ties = FALSE) |>
  select(teams, quarter, scores)
```
From the ggplot we can see that the data is pretty average, where all four quarters had somewhat equal scoring. When we look at the summarization table, we can see that the Celtics highest scoring quarter was Q3, Warriors was Q3, and Lakers was Q2. 

