---
title: "project_04"
format: html
toc: true
toc-depth: 5
toc-location: left
---

## Approach

I set up the files from https://spamassassin.apache.org/old/publiccorpus/?C=M;O=A into my project folder under ./data/ham and ./data/spam by downloading and extracting the most recently modified tar files. So, I'll use these for my spam and not-spam (bool) training data. 
 
I'll create a list for each folder, `list.files("/data/spam")` as spam_files and ``list.files("/data/ham")` as ham_files. I'll use a standard index for an ID (each row is an email), then I'll collapse each file into a single row of data in an str field called text. The tibble will also have a bool column for spam/ham where 0 will be ham and 1 will be spam. 

I'll then create a new df from this tibble where the words are unnested, so ID 1 will be n rows dependent on the word count, filter out common words using anti_join(stop_words), which is a tidytext function to remove common words. Then we will count for the next step of converting it into a data matrix for the prediction model. 

The data matrix will be a wide table where each word is given a column and each row is an email_id along with the associated counts per word per column. To do this I need to use library `tm` which is called the text mining package, which allows us to create a matrix from these words with relative ease. 

I asked AI what type of model would be best suited and it recommend the naiveBayes model, which requires the library `e1071`. So we will use that model with the dtm input and the emails df. 

Afterwards we can test the model by taking some of the older modified files of https://spamassassin.apache.org/old/publiccorpus/?C=M;O=A like the ones with hard_ham or maybe generate a csv of my own with AI generating 20-50 examples of spam/ham emails and see if the model gets it right. 

## Codebase

```{r}
library(tidyverse)
library(tidytext)
library(tm)
library(e1071)

spam_files <- list.files("./data/spam", full.names = TRUE) |>
  discard(~ basename(.x) == "0000.7b1b73cf36cf9dbc3d64e3f2ee2b91f1")

ham_files <- list.files("./data/ham", full.names = TRUE)

set.seed(123)

n_files <- min(length(spam_files), length(ham_files))

spam_files <- sample(spam_files, n_files)
ham_files  <- sample(ham_files, n_files)

length(spam_files)
length(ham_files)

```

```{r}

read_email <- function(path) {
  raw <- readLines(path, warn = FALSE, encoding = "latin1")
  
  subject <- raw[str_detect(raw, "^Subject:")] |>
    str_replace("^Subject:\\s*", "") |>
    paste(collapse = " ")
  
  blank_line <- which(raw == "")[1]
  
  body <- if (is.na(blank_line)) {
    raw
  } else {
    raw[(blank_line + 1):length(raw)]
  }
  
  paste(subject, paste(body, collapse = " "))
}

# build email tibble
spam_df <- tibble(
  file = spam_files,
  label = 1,
  text = map_chr(file, read_email)
)

ham_df <- tibble(
  file = ham_files,
  label = 0,
  text = map_chr(file, read_email)
)

emails <- bind_rows(spam_df, ham_df) |>
  mutate(email_id = row_number()) |>
  select(email_id, file, label, text)

emails <- emails[-1, ]

head(emails)

```

```{r}
email_words <- emails |>
  select(email_id, label, text) |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>
  filter(str_detect(word, "^[a-z]+$")) |>
  count(email_id, label, word, name = "count")

email_words

email_words |>
  distinct(email_id, label) |>
  mutate(label_name = if_else(label == 1, "spam", "ham")) |>
  count(label_name)

top_words <- email_words |>
  group_by(word) |>
  summarize(
    total = sum(count),
    docs = n_distinct(email_id),
    .groups = "drop"
  ) |>
  filter(docs >= 5) |>
  slice_max(total, n = 3000)

email_words_small <- email_words |>
  semi_join(top_words, by = "word")

head(email_words_small)

email_words_small |>
  distinct(email_id, label) |>
  mutate(label_name = if_else(label == 1, "spam", "ham")) |>
  count(label_name)

email_words2 <- email_words_small |>
  rename(.label = label)

email_dtm <- email_words2 |>
  pivot_wider(
    id_cols = c(email_id, .label),
    names_from = word,
    values_from = count,
    values_fill = 0
  )
```

```{r}
x <- email_dtm |>
  select(-email_id, -.label)

y <- as.factor(email_dtm$.label)

table(y)
```

```{r}
train_ids <- sample(seq_len(nrow(email_dtm)), 
                    size = 0.8 * nrow(email_dtm))

x_train <- x[train_ids, ]
x_test  <- x[-train_ids, ]

y_train <- y[train_ids]
y_test  <- y[-train_ids]
```

```{r}
model <- naiveBayes(x_train, y_train, laplace = 1)

pred <- predict(model, x_test)

table(predicted = pred, actual = y_test)

mean(pred == y_test)
```
0.725 accuracy is not that bad! I didn't use it against new data, but I did withhold 20% of the data within the corpus to be used for predictions. I think I can play around with the probabilities of the prediction, but to be honest, it took me a long time to just complete the project. 