---
title: "assignment 11"
format: html
toc: true
toc-depth: 5
toc-location: left
---

## Approach

### Prelim
I'm going to need to use the previous data which was created via assignment 3, so I brought over the code and the data. I modified the code to create an output called movie_ratings_fixed.csv that I will use for the recommender. 

```{r}
#| eval: false
library(tidyverse)

df <- read.csv("https://raw.githubusercontent.com/Siganz/CUNY_Assignments/refs/heads/main/607/assignment_11/movie_ratings.csv") |>
  select(name, title, rating)

# Global mean
global_mean <- mean(df$rating, na.rm = TRUE)

# Rater effects
s_name <- df |>
  summarize(rater_mean = mean(rating, na.rm = TRUE), .by = name) |>
  mutate(rater_effect = rater_mean - global_mean)

# Movie/item effects
s_title <- df |>
  summarize(item_mean = mean(rating, na.rm = TRUE), .by = title) |>
  mutate(item_effect = item_mean - global_mean)

# Fill original NA ratings
df2 <- df |>
  left_join(s_name, by = "name") |>
  left_join(s_title, by = "title") |>
  mutate(
    rating = if_else(
      is.na(rating),
      round(global_mean + rater_effect + item_effect),
      rating
    )
  )

# Add Shawn, who has no ratings
df3 <- df2 |>
  distinct(title, item_effect) |>
  mutate(
    name = "Shawn",
    rater_effect = 0,
    rating = round(global_mean + item_effect)
  ) |>
  select(name, title, rating)

# Combine fixed original data + Shawn rows
movie_ratings_fixed <- df2 |>
  select(name, title, rating) |>
  bind_rows(df3)

# Write CSV
write.csv(movie_ratings_fixed, "movie_ratings_fixed.csv", row.names = FALSE)

movie_ratings_fixed
```

### Main

I'll utilize `recommenderlab` and I'll do their UBCF method. The [github](https://github.com/mhahsler/recommenderlab) has an easy to follow usage. So I'll just use that and replace the data with the one developed in assignment 3.

```{r}
#| eval: false

df <- read.csv("https://raw.githubusercontent.com/Siganz/CUNY_Assignments/refs/heads/main/607/assignment_11/movie_ratings_fixed.csv")

train <- MovieLense100[1:300]
rec <- Recommender(train, method = "UBCF")
rec

pre <- predict(rec, MovieLense100[301:302], n = 5)
pre

scheme <- evaluationScheme(MovieLense100, method = "cross-validation", k = 10, given = -5,
    goodRating = 4)
scheme
```

## Codebase

```{r}
library(tidyverse)
library(recommenderlab)

df <- read.csv("https://raw.githubusercontent.com/Siganz/CUNY_Assignments/refs/heads/main/607/assignment_11/movie_ratings.csv")

rating_matrix <- df |>
  select(name, title, rating) |>
  pivot_wider(names_from = title, values_from = rating, values_fill = NA) |>
  column_to_rownames("name") |>
  as.matrix() |>
  as("realRatingMatrix")

rec <- Recommender(rating_matrix, method = "UBCF")

top_rec <- predict(rec, rating_matrix, n = 1)
as(top_rec, "list")

results <- tibble(
  user = rownames(rating_matrix),
  recommended = as(top_rec, "list")) |>
  unnest(recommended) |>
  rename(movie = recommended)

scheme <- evaluationScheme(rating_matrix, method = "cross-validation", k = 4, given = -2, goodRating = 4)

eval_results <- evaluate(scheme, method = "UBCF", type = "topNList", n = 1:5)

avg(eval_results)
plot(eval_results)
```

## Video

```{r}
df |> arrange(name, desc(rating))
results

```
The results show that the recommendations for each person are movies they didn't watch, which is accurate to the model.

```{r}
library(tidyverse)
library(recommenderlab)

df <- read.csv("https://raw.githubusercontent.com/Siganz/CUNY_Assignments/refs/heads/main/607/assignment_11/movie_ratings.csv")
```

Upon reviewing it makes little sense to do a recommendation model, probably should've done a missing value model instead. 
```{r}
rating_matrix <- df |>
  select(name, title, rating) |>
  pivot_wider(names_from = title, values_from = rating, values_fill = NA) |>
  column_to_rownames("name") |>
  as.matrix() |>
  as("realRatingMatrix")

rec <- Recommender(
  rating_matrix,
  method = "UBCF",
  parameter = list(method = "Cosine", nn = 3)
)

pred <- predict(
  rec,
  rating_matrix,
  type = "ratings"
)

original <- as(rating_matrix, "matrix")
predicted <- as(pred, "matrix")

missing <- is.na(original)

filled_missing_values <- tibble(
  name = rownames(original)[row(original)[missing]],
  title = colnames(original)[col(original)[missing]],
  predicted_rating = predicted[missing]
)

print(filled_missing_values)
```
```{r}
movie_means <- df |>
  group_by(title) |>
  summarise(
    mean_rating = mean(rating, na.rm = TRUE),
    n_ratings = sum(!is.na(rating)),
    .groups = "drop"
  ) |>
  arrange(desc(mean_rating))

print(movie_means)
```
Interesting to see that it goes beyond just calculating the mean. 