---
title: "assignment 11"
format: html
toc: true
toc-depth: 5
toc-location: left
---

## Approach

### Prelim
I'm going to need to use the previous data which was created via assignment 3, so I brought over the code and the data. I modified the code to create an output called movie_ratings_fixed.csv that I will use for the recommender. 

```{r}
#| eval: false
library(tidyverse)

df <- read.csv("https://raw.githubusercontent.com/Siganz/CUNY_Assignments/refs/heads/main/607/assignment_11/movie_ratings.csv") |>
  select(name, title, rating)

# Global mean
global_mean <- mean(df$rating, na.rm = TRUE)

# Rater effects
s_name <- df |>
  summarize(rater_mean = mean(rating, na.rm = TRUE), .by = name) |>
  mutate(rater_effect = rater_mean - global_mean)

# Movie/item effects
s_title <- df |>
  summarize(item_mean = mean(rating, na.rm = TRUE), .by = title) |>
  mutate(item_effect = item_mean - global_mean)

# Fill original NA ratings
df2 <- df |>
  left_join(s_name, by = "name") |>
  left_join(s_title, by = "title") |>
  mutate(
    rating = if_else(
      is.na(rating),
      round(global_mean + rater_effect + item_effect),
      rating
    )
  )

# Add Shawn, who has no ratings
df3 <- df2 |>
  distinct(title, item_effect) |>
  mutate(
    name = "Shawn",
    rater_effect = 0,
    rating = round(global_mean + item_effect)
  ) |>
  select(name, title, rating)

# Combine fixed original data + Shawn rows
movie_ratings_fixed <- df2 |>
  select(name, title, rating) |>
  bind_rows(df3)

# Write CSV
write.csv(movie_ratings_fixed, "movie_ratings_fixed.csv", row.names = FALSE)

movie_ratings_fixed
```

### Main

I'll utilize `recommenderlab` and I'll do their UBCF method. The [github](https://github.com/mhahsler/recommenderlab) has an easy to follow usage. So I'll just use that and replace the data with the one developed in assignment 3.

```{r}
#| eval: false

df <- read.csv("https://raw.githubusercontent.com/Siganz/CUNY_Assignments/refs/heads/main/607/assignment_11/movie_ratings_fixed.csv")

train <- MovieLense100[1:300]
rec <- Recommender(train, method = "UBCF")
rec

pre <- predict(rec, MovieLense100[301:302], n = 5)
pre

scheme <- evaluationScheme(MovieLense100, method = "cross-validation", k = 10, given = -5,
    goodRating = 4)
scheme
```