---
title: "assignment_10A"
format: html
toc: true
toc-depth: 5
toc-location: left
bibliography: references.bib
---

## Approach

### Set Up

```{r}
library(tidytext)
library(janeaustenr)
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
```
### Source Code

```{r}
tidy_books <- austen_books() %>%
  group_by(book) %>%
  mutate(
    linenumber = row_number(),
    chapter = cumsum(str_detect(
      text,
      regex("^chapter [\\divxlc]", ignore_case = TRUE)
    ))
  ) %>%
  ungroup() %>%
  unnest_tokens(word, text)

jane_austen_sentiment <- tidy_books %>%
  inner_join(get_sentiments("bing")) %>%
  count(book, index = linenumber %/% 80, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>% 
  mutate(sentiment = positive - negative)

ggplot(jane_austen_sentiment, aes(index, sentiment, fill = book)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~book, ncol = 2, scales = "free_x")
```

### Citations

[@silge2022; @yan2021]

### Implementation

I'll use AFINN instead of BING. I'll also try the gutenbergr package
```{r}
library(gutenbergr)
gutenberg_metadata |> filter(str_detect(author, "Vonnegut"))
vonnegut <- gutenberg_download(30240)
head(vonnegut)
```
I'll also use AFINN

```{r}
afinn <- get_sentiments("afinn")
```

## Codebase

```{r}
# transform Vonnegut file to get each line, chapter, word as it's own row. 
df_vonnegut <- vonnegut |>
  mutate(
    linenumber = row_number(),
    chapter = cumsum(str_detect(
      text,
      regex("^chapter [\\divxlc]", ignore_case = TRUE)
    ))
  ) |>
  ungroup() |>
  unnest_tokens(word, text)

# inner join the afinn data values per word
vonnegut_sentiment <- df_vonnegut |>
  inner_join(afinn, by = "word") |>
  mutate(index = linenumber %/% 80) |>
  group_by(index) |>
  summarise(
    sentiment = sum(value, na.rm = TRUE),       # net AFINN score
    n_words   = n()                              # words with sentiment scores
  ) |>
  ungroup()

vonnegut_sentiment |>
  ggplot(aes(x = index, y = sentiment)) +
  geom_col(fill = "darkred", alpha = 0.8) +
  geom_smooth(se = FALSE, color = "black") +
  labs(title = "Sentiment Progression in the Vonnegut Book (AFINN)",
       x = "Section (every 80 lines)",
       y = "Net Sentiment Score") +
  theme_minimal()
```

### Summary
This is a tiny book / short story from Vonnegut that is dark and satirical. Using AFINN, we see the darker sentiment in the middle when the family conflict gets worse and everything becomes more hostile and tense. Then we see more levity near the end, when Lou and Em decide they actually like being in prison because it finally gives them peace and privacy. That shift fits the story well, so it is interesting that AFINN is able to follow the basic emotional arc even if it cannot fully understand the satire.

