---
title: "assignment_10B"
format: html
toc: true
toc-depth: 5
toc-location: left
---

## Approach

### Brainstorm 

```{r}
library(tidyverse)
library(jsonlite)

df <- fromJSON("https://api.nobelprize.org/2.1/nobelPrizes")
glimpse(df)
```

Lot of nested dataframes and lists. Seems better to read the [documentation](https://app.swaggerhub.com/apis/NobelMedia/NobelMasterData/2.0) where it's better to create an https request. Unless the request is to use `https://api.nobelprize.org/2.1/nobelPrizes`? 

### Questions

After understanding the data, I can come up with the following questions: 

- Did Friedrich Nietzsche ever win a Nobel Prize?
- What was Betrand Russell's motivation for winning a prize?
- Who won the Nobel Prize in Literature in 2010-2020?
- What is the overall ratio of male to female Nobel Prize laureates? Among the five countries with the most birthed laureates, what is the male-to-female ratio in each country?

#### Nietzsche

Bool response for the query: `http://api.nobelprize.org/2.0/laureates?name=Friedrich%20Nietzsche`

#### Betrand Russell

Return motivation columns from the query: `http://api.nobelprize.org/2.0/laureates?name=Betrand%20Russell`

#### Nobel Prize Literature 2010-2020

`http://api.nobelprize.org/2.0/laureates?nobelPrizeYear=2010&yearTo=2020&nobelPrizeCategory=lit`

#### Ratio of male to Female Nobel Prize laureates and country outliers. 

We would use the full laureates dataset to calculate the overall male-to-female ratio among Nobel Prize laureates. Then we would group laureates by birthContinent, select the five continents with the most laureates, and calculate the ratio for each continent.

`http://api.nobelprize.org/2.0/laureates`

## Codeurl

### packages
```{r}
library(tidyverse)
library(jsonlite)
library(tidyr)
library(stringr)
```

### set up
```{r}
url <- "https://api.nobelprize.org/2.1"

# calling function makes it easier to answer the different questions
get_api <- function(path) {
  fromJSON(path, flatten = TRUE)
}
```

### Q1: Did Friedrich Nietzsche ever win a Nobel Prize?
```{r}
nietzsche <- get_api(
  paste0(url, "/laureates?name=Friedrich%20Nietzsche")
)

if (length(nietzsche$laureates) > 0) "Yes" else "No"
```

### Q2: What was he motivation for Bertrand Russell’s Nobel Prize?
```{r}
russell <- get_api(
  paste0(url, "/laureates?name=Bertrand%20Russell")
)

russell$laureates|>
  select(fullName.en, nobelPrizes) |>
  unnest(cols = nobelPrizes) |>
  select(fullName.en, awardYear, category.en, motivation.en) |>
  pull(motivation.en)

```

### Q3: Who won the Nobel Prize in Literature from 2010 to 2020?
```{r}
lit <- get_api(
  paste0(url, "/nobelPrizes?nobelPrizeCategory=literature&awardYear=2010&yearTo=2020")
)

lit_winners <- lit$nobelPrizes|>
  select(awardYear, laureates)|>
  unnest(cols = laureates) |>
  transmute(
    year = awardYear,
    winner = coalesce(knownName.en, fullName.en)
  )

lit_winners
```

### Q4: Overall male-to-female ratio, and the ratio in the 5 birth countries with the most laureates
```{r}

all_laureates <- get_api(
  paste0(url, "/laureates?limit=2000")
)

laureates_df <- all_laureates$laureates|>
  transmute(
    name = coalesce(knownName.en, fullName.en),
    gender,
    birth_country = birth.place.country.en
  )

overall_ratio <- laureates_df |>
  filter(gender %in% c("male", "female")) |>
  count(gender) |>
  pivot_wider(names_from = gender, values_from = n) |>
  mutate(male_to_female_ratio = male / female)

overall_ratio

top5_countries <- laureates_df |>
  filter(!is.na(birth_country)) |>
  count(birth_country, sort = TRUE) |>
  slice_head(n = 5)

top5_countries

country_ratios <- laureates_df |>
  filter(birth_country %in% top5_countries$birth_country,
         gender %in% c("male", "female")) |>
  count(birth_country, gender) |>
  pivot_wider(
    names_from = gender,
    values_from = n,
    values_fill = 0
  )

if (!"male" %in% names(country_ratios)) {
  country_ratios$male <- 0L
}

if (!"female" %in% names(country_ratios)) {
  country_ratios$female <- 0L
}

country_ratios <- country_ratios |>
  mutate(
    total = male + female,
    male_to_female_ratio = if_else(female == 0, NA_real_, male / female),
    ratio_text = if_else(
      female == 0,
      "No women; 100% male",
      paste0(round(male / female, 2), ":1")
    )
  ) |>
  arrange(desc(total))

country_ratios |>
  arrange(male_to_female_ratio)

country_ratios |>
  select(birth_country, male, female) |>
  pivot_longer(cols = c(male, female),
               names_to = "gender",
               values_to = "count") |>
  ggplot(aes(x = birth_country, y = count, fill = gender)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Male and Female Nobel Laureates by Birth Country",
    x = "Birth Country",
    y = "Count"
  ) +
  theme_minimal()
```

France has the strongest female representation relative to men, while the United States has the highest raw counts of both female and male laureates. In our sample, countries such as Japan and the U.K. had no female laureates. These results are based on birth_country, not the country where a laureate later lived or grew up.