---
title: "assignment_07"
format: html
toc: true
toc-depth: 5
toc-location: left
---

## Approach

I'll ask, were the top 10 viewed stories in the last 30 days?

This is an example of their api: `GET https://api.nytimes.com/svc/mostpopular/v2/viewed/30.json?api-key=YOUR_KEY`

### Schema

#### Core Fields

-   **url** *(string)* — Article URL\
-   **adx_keywords** *(string)* — Semicolon-separated keywords\
-   **section** *(string)* — Section (e.g., "Science")\
-   **byline** *(string)* — Author line (e.g., "By John Doe")\
-   **type** *(string)* — Asset type (e.g., "Article", "Interactive")\
-   **title** *(string)* — Headline\
-   **abstract** *(string)* — Summary\
-   **published_date** *(string)* — Format: YYYY-MM-DD\
-   **source** *(string)* — Publisher\
-   **id** *(integer)* — Asset ID\
-   **asset_id** *(integer)* — Same as id\

------------------------------------------------------------------------

#### Facets (Categorization)

-   **des_facet** *(array*<string>) — Topics / descriptions\
-   **org_facet** *(array*<string>) — Organizations\
-   **per_facet** *(array*<string>) — People\
-   **geo_facet** *(array*<string>) — Locations

------------------------------------------------------------------------

#### Media (array)

Each item:

##### Media

-   **type** *(string)* — e.g., "image"\
-   **subtype** *(string)* — e.g., "photo"\
-   **caption** *(string)* — Description\
-   **copyright** *(string)* — Credit\
-   **approved_for_syndication** *(boolean)*\
-   **media-metadata** *(array)*

##### MediaMetadata

-   **uri** *(string)* — Unique identifier\
-   **url** *(string)* — Image URL\
-   **width** *(integer)*\
-   **height** *(integer)*

I'll save my api-key into an r env that will be kept secret and not uploaded to github.

Then I'll transform this into a tibble, view the data, create tibble, select the columns to keep, then do a summarization table, then slice_max with n = 10. 

## Codebase

### Initialization
```{r}
library(tidyverse)
library(jsonlite)
library(ggthemes)

readRenviron(".Renviron")
api_key <- Sys.getenv("API_KEY")

days <- 30

url <- paste0("https://api.nytimes.com/svc/mostpopular/v2/viewed/", days, ".json?api-key=", api_key)
res <- fromJSON(url)
glimpse(res)
```

### Reading the Results

The fromJSON function returned a list of four, of which $results is what I really want. 

```{r}
nyt_top_articles <- res$results
glimpse(nyt_top_articles)
```

Interesting that the Schema does not include a ranking system for the top 20 pulls. The API also doesn't allow me to get top #, it always returns 20. I'll just create a sub-table of the results and I think it's safe to assume that the top 20 articles that get returned are already in order of popularity, so I'll create a new field called rank and have it valued as row number. 

```{r}
nyt <- nyt_top_articles|>
  select(id, title, section, subsection, abstract, adx_keywords) |>
  mutate(rank = row_number()) |>
  relocate(rank)
```

### Quick Analysis

We can view the top 10 articles via:

```{r}
nyt |> 
  select(rank, title)
```
We can see that the # 1 article is the Robert S. Mueller Death, which was a big headline at the time. Out of curiosity I also did a plot about section/categories, because I noticed that most of these articles are political. 

```{r}
nyt |> 
  count(section) |>
  ggplot(aes(x = reorder(section, n), y = n / sum(n), fill = section)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = 'NYT: Top Articles in The Last 30 Days by Section',
    subtitle = "Data was derived from NYT 'most viewed' api. 20 articles are the standard return value.",
    y = 'percent',
    x = 'category'
  ) +
  theme_fivethirtyeight() +
  theme(legend.position = "none")

```

From our quick analysis we can see that over 60% of the popular articles were from their U.S. section.